---
output: "github_document"
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# LSOApop

<!-- badges: start -->
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
<!-- badges: end -->

The goal of `LSOApop` is to make population estimates for **Lower layer Super Output Areas (LSOA)** and their **Indecies of Multiple Deprivation (IMD)** easily available in R.
Population estimates are broken down by age (0 to 90+) and gender (female/male).
A transparent description of the data processing and information about the original sources are available in this GitHub, see `"data-raw/imd.R` and `"data-raw/lsoa.R`.
The current version of the package only includes LSOA population estimates and IMD scores for the year 2019 for England. 
Because we store quite a lot in this package it currently relatively large (~7mb) compared to other packages.

## Installation

You can install the current version of `LSOApop` from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("nhs-r-community/LSOApop")
```

## Example

```{r}
# Load the package
library(LSOApop)
```

### Lower layer Super Output Areas (LSOA)

The LSOA population estimates are available in the dataset `lsoa`:

```{r}
# Show the first 6 rows of the dataset
# For further information about this dataset see the help file: help(lsoa)
head(lsoa)
```
### Indecies of Multiple Deprivation (IMD)

The IMD scores (raw scores and ranked deciles) and available in the dataset `imd`:

```{r}
# Show the first 6 rows of the dataset
# For further information about this dataset see the help file: help(imd)
head(imd)
```


## Sources of Data

The original source of the data provided in this R package is available [here](https://www.ons.gov.uk/peoplepopulationandcommunity/populationandmigration/populationestimates/datasets/lowersuperoutputareamidyearpopulationestimates) and licenced under the [Open Government Licence v3.0](http://www.nationalarchives.gov.uk/doc/open-government-licence/version/3/).


## Code of Conduct
  
Please note that the `LSOApop` project is released with a [Contributor Code of Conduct](https://contributor-covenant.org/version/2/0/CODE_OF_CONDUCT.html). 
By contributing to this project, you agree to abide by its terms.
