---
title: "How to calculate 'local' IMD ranks"
output: rmarkdown::html_vignette
bibliography: "references.bib"
link-citations: TRUE
vignette: >
  %\VignetteIndexEntry{calc-local-imd-ranks}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, message=FALSE, warning=FALSE}
# Load R packages ----
library(NHSRpopulation)
library(dplyr)
library(ggplot2)
library(tidyr)
library(patchwork)
```

[Indices of Multiple Deprivation (IMD)](https://www.gov.uk/government/statistics/english-indices-of-deprivation-2019) describe the deprivation of [Lower layer Super Output Areas (LSOAs)](https://data.gov.uk/dataset/c481f2d3-91fc-4767-ae10-2efdf6d58996/lower-layer-super-output-areas-lsoas) with scores ranging from `r min(imd$imd_score)` to `r max(imd$imd_score)` for the year 2019.

In this example we are looking at the IMD scores and decile ranks for Local 
Authorities (LAs) in Nottingham City and Nottinghamshire County. 
First, we define the corresponding Local Authority Codes to make it easier to 
work with only these regions.

```{r define-las}
# Nottingham City Local Authority and Nottinghamshire County Council LA areas
# often have their data combined, particularly for the Provider
# Trusts like Nottinghamshire Healthcare NHS Foundation Trust.

# Define code for Nottingham City LA
la_code_nott_city <- c("E06000018")

# Define codes for Nottinghamshire County Council LA districts (further details
# can be found here
# https://cdu-data-science-team.github.io/team-blog/posts/2021-05-14-index-of-multiple-deprivation/#imd-in-sql)
la_code_notts_county <- c(
  "E06000018", "E07000170", "E07000171", "E07000172",
  "E07000173", "E07000174", "E07000175", "E07000176"
)

# Combine all districts into one object
la_code_nott_city_notts_county <- c(la_code_nott_city, la_code_notts_county)
```

Lower IMD scores indicate 'less deprivation' areas while higher IMD scores 
indicate 'more deprivation'.
Figure 1 shows the raw scores for each of the different LAs with colour codings 
for the IMD decile ranks that were calculated using data from all LAs in 
England.

```{r fig.width=7, fig.height=5}
imd %>%
  mutate(imd_decile_factor = factor(imd_decile,
    levels = 1:10,
    labels = c("1 - Most deprived", 2:9, "10 - Least deprived")
  )) %>%
  filter(la_code %in% la_code_nott_city_notts_county) %>%
  ggplot(aes(x = imd_score, y = la_name, fill = imd_decile_factor)) +
  geom_jitter(shape = 21, height = .2, alpha = 0.8, size = 2) +
  scale_fill_viridis_d(direction = -1) +
  labs(
    x = "IMD score",
    fill = "ENG IMD decile",
    size = "ENG IMD decile",
    y = NULL
  ) +
  ggtitle("Fig. 1. IMD scores for LSOAs in Nott City and Notts County") +
  scale_x_reverse()
```

<!-- The distribution of the IMD deciles calculated on data from the entire 
country can be unequal within some Local Authorities (LAs) or groups of LAs. -->
<!-- This example illustrates this for the LA Nottingham 'City' and the LAs in 
Nottinghamshire 'County'. -->

## IMD ranks

As mentioned above, these raw IMD scores can be grouped based on their rank, 
usually 10 groups (deciles) are calculated based on the IMD scores of the entire 
country. 
For the decile scores, lower deciles are representing 'more deprived' areas 
while higher deciles represent 'less deprived'. This can be confusing because 
IMD scores and IMD deciles are scored in opposite directions.

### England IMD deciles

Figure 2.1 below illustrates that the distribution of IMD deciles calculated 
based on the whole country (England) can be relatively unequal when comparing 
individual LAs (e.g., Nottingham (City) versus Rushcliffe). 
In some cases using the deciles calculated based on IMD scores for the entire 
country may miss important regional differences and it could be useful to 
recalculate IMD deciles within Local Authorities. 

Whilst the @MinistryofHousingCommunitiesLocalGovernment2019 has given some 
guidance on how to calculate and analyse IMD scores, there are no clear examples 
or guidance when and how to calculate local IMD ranks.

```{r fig.width=7, fig.height=5}
# Plot deciles ----
imd %>%
  mutate(imd_decile = factor(imd_decile,
    levels = 1:10,
    labels = c("1 - Most deprived", 2:9, "10 - Least deprived")
  )) %>%
  filter(la_code %in% la_code_nott_city_notts_county) %>%
  ggplot(aes(factor(imd_decile), fill = factor(imd_decile))) +
  geom_bar() +
  scale_fill_viridis_d() +
  scale_x_discrete(labels = 1:10) +
  labs(x = "IMD decile", y = "Count", fill = "IMD decile") +
  ggtitle("Fig. 2.1. Combined ENG IMD deciles for LSOAs in Nott City and Notts County")
```


```{r fig.width=7, fig.height=7}
imd %>%
  mutate(imd_decile = factor(imd_decile,
    levels = 1:10,
    labels = c("1 - Most deprived", 2:9, "10 - Least deprived")
  )) %>%
  filter(la_code %in% la_code_nott_city_notts_county) %>%
  ggplot(aes(factor(imd_decile), fill = factor(imd_decile))) +
  geom_bar() +
  scale_fill_viridis_d() +
  labs(x = "IMD decile", y = "Count", fill = "IMD decile") +
  facet_wrap(~la_name, ncol = 2) +
  scale_x_discrete(labels = 1:10) +
  ggtitle("Fig. 2.2. Individual ENG IMD deciles for LSOAs in Nott City and Notts County")
```

### Calculate 'local' IMD deciles within Local Authorities

The following example illustrates how local IMD ranks can be calculated within 
a selected group of LAs (see first code chunk where a list of LA codes is 
defined: `la_code_nott_city_notts_county`). 

* To recalculate IMD deciles for Local Authorities the data needs to be grouped 
by a defined group of Local Authorities (see code below, the variable `la_group` 
gets calculated for this example) and year of the IMD score 
('`group_by(la_code, imd_year)`').

* Next, calculate deciles using the function '`dplyr::ntile()`' by specifying 
'`n = 10`'.

* To calculate quintiles (5 groups) the argument would need to be changed to 
`n = 5`.

To take into account that IMD scores and IMD ranks are scored in the opposite 
direction, we need to add a minus before the IMD score: 
`ntile(-imd_score, n = 10)`.

```{r}
# Calculate IMD decile ranks (n = 10) for local authorities ----
imd_by_la <- imd %>%
  mutate(la_groups = case_when(
    la_code %in% la_code_nott_city_notts_county ~ "notts_city_county"
  )) %>%
  filter(la_groups == "notts_city_county") %>%
  group_by(la_groups, imd_year) %>%
  mutate(imd_decile_la = ntile(-imd_score, n = 10))
```

### Local IMD deciles for Nott City and Notts County

As expected, Figure 3.1 shows an equal distribution of recalculated 'local' IMD 
deciles for the LAs Nottingham City and Nottinghamshire County.

```{r fig.width=7, fig.height=5}
# Plot ENG and NOTTS deciles ----
imd_by_la %>%
  mutate(imd_decile_la = factor(imd_decile_la,
    levels = 1:10,
    labels = c("1 - Most deprived", 2:9, "10 - Least deprived")
  )) %>%
  filter(la_code %in% la_code_nott_city_notts_county) %>%
  ggplot(aes(factor(imd_decile_la), fill = factor(imd_decile_la))) +
  geom_bar() +
  scale_fill_viridis_d() +
  labs(x = "IMD decile", y = "Count", fill = "IMD decile") +
  scale_x_discrete(labels = 1:10) +
  ggtitle("Fig. 3.1. Combined 'local' IMD deciles for LSOAs in Nott City and Notts County")
```

Figure 3.2 shows the recalculated 'local' IMD ranks for each individual 
selected LA.
There are sill differences in the distributions of 'local' IMD ranks across the 
different LAs.
However, these rankings were recalculated using only the IMD scores of the LAs 
that were of interest and may therefore be a better presentation of the 
variation across these LAs.

```{r fig.width=7, fig.height=7}
# Plot ENG and NOTTS deciles ----
imd_by_la %>%
  mutate(imd_decile_la = factor(imd_decile_la,
    levels = 1:10,
    labels = c("1 - Most deprived", 2:9, "10 - Least deprived")
  )) %>%
  filter(la_code %in% la_code_nott_city_notts_county) %>%
  ggplot(aes(factor(imd_decile_la), fill = factor(imd_decile_la))) +
  geom_bar() +
  scale_fill_viridis_d() +
  labs(x = "IMD decile", y = "Count", fill = "IMD decile") +
  facet_wrap(~la_name, ncol = 2) +
  scale_x_discrete(labels = 1:10) +
  ggtitle("Fig. 3.2. Individual 'local' IMD deciles for LSOAs in Nott City and Notts County")
```

## Compare distributions of IMD deciles 

The changes in IMD decile ranks can be seen when comparing the bar plots shown 
above (e.g., Fig 2.2 vs. Fig 3.2) -- but they can be hard to spot.

Changes in the distribution of IMD deciles (see density plots in Figure 4) 
illustrate the effect of recalculating 'local' IMD scores a bit clearer.

* The plot on the left with the title '**ENG IMD**' shows the distribution of 
IMD deciles when all IMD scores in England were used to calculate the decile 
ranks.

* The plot on the right (**Local IMD**) shows the distribution of IMD deciles 
when only the IMD scores of the selected LAs were used.

The effect of recalculating local IMD scores can be seen in changes in the shape 
of the density distribution (e.g., narrow shape vs flat shape, changes of the 
peak on the x axis) for the individual IMD deciles.


```{r fig.width=7, fig.height=3.5}
imd_by_la %>%
  pivot_longer(
    cols = c(imd_decile, imd_decile_la),
    names_to = "imd_method", values_to = "imd_value"
  ) %>%
  mutate(imd_value = factor(imd_value,
    levels = 1:10,
    labels = c("1 - Most deprived", 2:9, "10 - Least deprived")
  )) %>%
  filter(la_code %in% la_code_nott_city_notts_county) %>%
  ggplot(aes(x = imd_score, fill = factor(imd_value))) +
  scale_fill_viridis_d(direction = 1) +
  labs(x = "IMD score", y = NULL, fill = "IMD decile") +
  geom_density(alpha = .6, adjust = 2) +
  facet_wrap(~ factor(imd_method,
    levels = c(
      "imd_decile",
      "imd_decile_la"
    ),
    labels = c(
      "END IMD",
      "Local IMD"
    )
  )) +
  ggtitle("Fig. 4. Distributions of ENG and local IMD deciles")
```


## Additional resources

A blog post by @Turner2021IMD also discusses how to calculate local IMD scores 
illustrating examples in SQL and R.

## Notes for anyone contributing to the html and vignettes 

The pages built here https://nhs-r-community.github.io/NHSRpopulation/ have 
been done using the package [{pkgdown}](https://pkgdown.r-lib.org/). 
In order to rebuild the pages after any changes use `pkgdown::build_site()` in 
the console. There will be a number of changes to the supporting files in the 
folder `docs/` which all then need to be committed.

## References
<!-- References are compiled automatically from text in the main areas preceded 
by the @ sign like @Turner2021IMD. This note is hidden from the 
published sites. -->
