---
title: "Getting started"
output: rmarkdown::html_vignette
bibliography: "references.bib"
link-citations: TRUE
vignette: >
  %\VignetteIndexEntry{get-started}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  echo = TRUE,
  eval = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

### Indices of Multiple Deprivation (IMD)

To get the IMD scores (raw scores and ranked deciles) for a dataset run the following code to generate some random example postcodes:

```{r}
library(purrr)
library(tibble)
library(PostcodesioR)
library(NHSRpopulation)

postcodes <- purrr::map_chr(
  1:10,
  .f = ~PostcodesioR::random_postcode() |> 
    purrr::pluck(1) 
) 

tibble_postcodes <- postcodes |> 
  tibble::as_tibble()
```

Then, using the `get_imd()` function for a vector (returning just the first five columns):

```{r}
NHSRpopulation::get_imd(postcodes) |> 
  dplyr::select(1:5)
```

Or with a data frame (returning just the first five columns):

```{r}
NHSRpopulation::get_imd(tibble_postcodes$value) |> 
  dplyr::select(1:5)
```

This function can be used to fix missing postcodes as some are terminated or are invalid:

```{r}
postcodes <- c("HD1 2UT", "HD1 2UU", "HD1 2UV")

NHSRpopulation::get_imd(postcodes) |> 
  dplyr::select(1:5)

```

Currently, although the postcode is fixed with the column `new_postcode` the IMD is not overwritten.

## Lower Super Output area (LSOA)

To return the `IMD`, `imd_decile` and `imd_quintile` for LSOAs this can be as a vector:

```{r}
# Example LSOAs from each England Decile group
lsoa_imd <- c("E01000002",
              "E01000001",
              "E01000117",
              "E01000119",
              "E01000069",
              "E01000070",
              "E01000066",
              "E01000005",
              "E01000008",
              "E01000048")

NHSRpopulation::get_lsoa(lsoa_imd) |> 
  head(10) # first 10 rows
```

Or from a data frame:

```{r}

tibble_lsoa_imd <- lsoa_imd |> 
  tibble::as_tibble()

NHSRpopulation::get_lsoa(tibble_lsoa_imd$value) |> 
  head(10)

```

The functions return everything in those LSOAs and if you would like to return some random postcodes from each decile:

```{r}
NHSRpopulation::get_lsoa(lsoa_imd, return = "random")
```

Or just the first postcode that appears in each decile:

```{r}
NHSRpopulation::get_lsoa(lsoa_imd, return = "first")
```
