---
title: "Getting started with NHSRpopulation"
output: rmarkdown::html_vignette
link-citations: TRUE
vignette: >
  %\VignetteIndexEntry{Getting started with NHSRpopulation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  echo = TRUE,
  eval = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

### Indices of Multiple Deprivation (IMD)

To get the IMD scores (raw scores and ranked deciles) for a dataset run the 
following code to generate some random example postcodes:

```{r, eval=TRUE}
library(tibble)
library(NHSRpopulation)

postcodes <- c("HD1 2UT", "HD1 2UU", "HD1 2UV")

tibble_postcodes <- postcodes |>
  tibble::as_tibble() |>
  dplyr::rename(postcode = value)

tibble_postcodes
```

Then, using the `get_data()` function for a vector:

```{r,}
# Execution halted
NHSRpopulation::get_data(postcodes) |>
  dplyr::select(
    postcode,
    new_postcode,
    result_type,
    lsoa_code
  )
```

Or with a data frame:

```{r}
NHSRpopulation::get_data(tibble_postcodes) |>
  dplyr::select(
    postcode,
    new_postcode,
    result_type,
    lsoa_code
  )
```

Note that this function uses the {NHSRpostcodetools} package to offer the 
opportunity to fix postcodes which are terminated or are incorrect. 
This is default and appears in the column `new_postcode` and does not overwrite
the original `postcode` column.
Switching off this automatic fix can be done with the code and will accept both
vectors and data frames:

```{r}
NHSRpopulation::get_data(tibble_postcodes, fix_invalid = FALSE) |>
  dplyr::select(
    postcode,
    new_postcode,
    result_type,
    lsoa_code
  )
```

# Index of Multiple Deprivation


```{r, eval=TRUE}
# Note that the third LSOA in this list is incorrect on purpose
imd <- c("E01011107", "E01011229", "E01002")

tibble_imd <- imd |>
  tibble::as_tibble() |>
  dplyr::rename(lsoa11 = value)

tibble_imd
```

Using the same function but with a parameter/argument to return IMD data:

```{r}
NHSRpopulation::get_data(tibble_imd, url_type = "imd") |>
  dplyr::select(
    lsoa11,
    imd_rank,
    imd_decile,
    imd_score
  )
```

Data can be either vectors or data frames.

No corrections are made to incorrect LSOA codes and the current join to the API
is on `lsoa11` data.

# Column names

Where data frames are used the expectation of the functions is that postcodes
will be in a column called `postcode` and IMD will be from `lsoa11`, however, 
this can be overwritten:

```{r}
# Create datasets
pcs_tb <- dplyr::tibble(
  pcs = postcodes
)

pcs_tb  

lsoa_tb <- dplyr::tibble(
  lower_soa = imd
)

lsoa_tb
```

The argument/parameter `column = ` can be used to set the column name:

```{r}
NHSRpopulation::get_data(pcs_tb,
  column = "pcs"
)
```

```{r}
NHSRpopulation::get_data(lsoa_tb,
  column = "lower_soa"
)
```

```{r}
NHSRpopulation::get_data(different_names,
  # url_type is set to postcodes as default
  column = "pcs"
)
```

